The effect of prenatal environment on brain metabolic function and perception in a lizard

### 2025-10-13 ms.qmd
updates

Cognition encompasses the ways in which animals acquire, process, and store information, enabling perception, learning, memory, and decision-making [@shettleworth]. It is essential for survival and reproduction, allowing individuals to adapt to changing environments [@dukas_evolutionary_2004]. However, cognitive abilities can vary considerably between individuals, with differences arising from genetic factors, environmental conditions, or a combination of both [@sakata_neural_2000; @dukas_evolutionary_2004]. The prenatal environment, in particular, plays a critical role in shaping brain development and cognition across a wide range of species [@sakata_neural_2000; @zhu_prenatal_2004; @crino_corticosterone_2014-learn; @amiel_effects_2017]. However, the impact of early-life environments has primarily been studied in the context of learning and memory [@zhu_prenatal_2004; @crino_corticosterone_2014-learn; @bebus_associative_2016; @abayarathna_effects_2020], neglecting how other important cognitive domains might be affected [but see @burger_antipredator_1998; @vila_pouca_quantity_2019]. For example, perception is an often overlooked, but fundamental cognitive process [@shettleworth]. Perceiving different stimuli is essential for behaviours critical to survival and reproductive succes such as locating food, avoiding predators, or interacting with conspecifics [@hairston1982fish; @pyke1984optimal; @burger_antipredator_1998; @desfilis2003stimulus; @recio2023conspecific]. Therefore, understanding how prenatal conditions affect the ability to detect prey can be important to understanding the broader consequences of early-life environments on fitness.

Prenatal environments influence cognitive abilities because the brain is particularly sensitive to environmental conditions during early stages of development [@zhu_prenatal_2004]. Early-life conditions can shape cognition by altering gene expression [@zhou2020effects], neurotransmitter production and neuroendocrine pathways [@amani2021perinatal], or brain structure and function [@amiel_effects_2017]. Additionally, developmental environments can have long-lasting effects on energy production and oxidative stress in the brain, which may be a key mechanism influencing cognitive function [@siegel1994basic; @du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018]. First, given the high energetic demands of cognitive processes [@mcnay_decreases_2000; @mann_coupling_2021; @alexandrov_neuronal_2022], cognitive performance is intrinsically linked to efficient mitochondrial respiration. Likewise, high cognitive abilities are typically associated with increased neuron density and functionality [@lefebvre_taxonomic_2011; @amiel_effects_2017], both of which can be impaired by excessive reactive oxygen species (ROS) production and oxidative stress [@finkel_oxidants_2000; @zhu_prenatal_2004; @du_dynamic_2009; @gong_chronic_2011; @hoffmann_mitochondrion_2018]. For instance, Hara et al. (2014) found that performance in visuospatial working-memory tasks is negatively correlated to the number of round-shaped mitochondria-associated with higher oxidative stress-per presynaptic terminal. Findings from different studies suggest that mitocondrial activity can be essential in shaping cognitive abilities. Thus, understanding how early-life conditions affect mitochondrial physiology may provide insights into the mechanisms underlying cognitive performance.

Mitochondria are highly responsive to environmental conditions during early development, with potential short-term and long-lasting consequences [reviewed in @gyllenhammer2020developmental]. For example, environmental conditions or hormone exposure can cause immediate disruptions to metabolic function, triggering cascading effects on oxidative stress that may persist later in life [reviewed in @gyllenhammer2020developmental]. Among the range of environmental factors that can shape mitochondrial physiology, some exert particularly strong and consistent effects. Stress-related hormones and temperature are particularly relevant for mitochondrial function due to their strong effects on metabolic processes [@sapolsky_how_2000; @stier2022experimental; @crino2024eggs]. Vertebrates respond to stressful situations through the activation of physiological processes including the production of  glucocorticoids (GCs), which are also important metabolic regulators [@sapolsky_how_2000]. In endothermic vertebrates, temperature plays a similar role in regulating metabolic activity [@stier2022experimental; @crino2024eggs]. In fact, heightened GC levels or changes in the thermal environment during early development have profound and sustained effects on mitochondrial respiration efficiency or oxidative stress [@zhu_prenatal_2004; @song2009prenatal; @costantini2011meta; @haussmann_embryonic_2012; @treidel2016temperature; @stier2022experimental; @crino2024eggs]. However, how the early environment influences mitochondrial physiology, and its associated cognitive consequences remains largely unknown outside of mammals [see @chaudhari2022early for a review]. In addition, whether and how these interactions affect perceptual abilities, such as prey detection, is still largely unexplored.

@@

#### Animal husbandry  
*Breeding colony* -- The lizards tested came from a breeding colony established in the laboratory in 2019. The colony consisted of 270 adults of *L. delicata* housed in plastic containers (41.5 L x 30.5 W x 21 H cm) with six lizards (two males and four females) per enclosure. Enclosures were provided with non-stick matting, shelter, and several small water dishes. Water was given daily, and lizards were fed approx. 40 mid-size crickets (_Acheta domestica_) per enclosure three days a week. Crickets were dusted with calcium weekly and multivitamin and calcium biweekly. Room temperatures were set to 22-24 ºC, but to ensure a temperature gradient, we provided the enclosures with a heat chord and a heat lamp following a 12 h light:12 h dark cycle keeping warm side of enclosures is usually at 34 ºC.

*Egg collection and incubation* -- Between mid-October 2022 and the end of February 2023, we provide females with a place to lay the eggs by placing a small box (12.5 L x 8.3 W x 5 H cm) with moist vermiculite in one side of the communal enclosures. These boxes were checked three days a week for eggs. After collection, we measured the length and width of eggs with a digital caliper to the nearest 0.1 mm and weighed them with a digital scale (± 0.001 g). Eggs were treated with either CORT or a vehicle control (see CORT and temperature manipulation below) and placed in individual cups (80 mL) with moist vermiculite (12 parts water to 4 parts vermiculite). Cups were covered with cling wrap to retain moisture and left at one of two different incubation temperatures (see CORT and temperature manipulation below) until hatching.

*Hatchlings* -- Incubators were checked three times a week for hatchlings. We measured and weighted lizards immediately after hatching. We used a ruler to measure snout-vent length (SVL) and tail length (TL) to the nearest mm, and a digital scale to measure mass (± 0.001 g). We then placed hatchlings in individual enclosures (18.7L x 13.2W x 6.3H cm) and provided them with nonstick matting and a small water dish until the beginning of the experiment. During this period, lizards were given water daily and received 3-6 small *A. domestica* crickets three times a week. All care otherwise follows similar protocols to adults (see above).

@@

To test the interactive effects of CORT and incubation temperature, we manipulated CORT concentrations in eggs and incubated them under one of two temperature regimes (see @fig-Methods A). We used a partial split clutch design where eggs from a given clutch were distributed equally across the four treatments when clutch sizes were larger than four and randomly across treatments when less than four. Eggs were topically supplied with either: a) 5 µL of crystalline corticosterone (Sigma, Cat. No. C2505) dissolved in 100% ethanol at a final 10 pg CORT/mL concentration (CORT treatment), or b) an equal volume of 100% Ethanol (Control treatment). We selected doses based on our previous study where CORT treatment increased mean yolk CORT levels by approximately 2 standard deviations above the mean natural concentration [@crino2024eggs]. Eggs were then incubated at either cold  (23 ± 3 ºC) or hot (28 ± 3 ºC) incubation temperatures in a 2x2 factorial design (Fig. 1 A). These temperatures are within the natural limits in _L. delicata_ [@cheetham2011embryonic].

>>>>

Cognition encompasses the ways in which animals acquire, process, and store information, enabling perception, learning, memory, and decision-making [@shettleworth]. It is essential for survival and reproduction, allowing individuals to adapt to changing environments [@dukas_evolutionary_2004]. However, cognitive abilities can vary considerably between individuals, with differences arising from genetic factors, environmental conditions, or a combination [@sakata_neural_2000; @dukas_evolutionary_2004]. The prenatal environment, in particular, plays a critical role in shaping brain development and cognition across a wide range of species [@sakata_neural_2000; @zhu_prenatal_2004; @crino_corticosterone_2014-learn; @amiel_effects_2017]. However, the impact of early-life environments has primarily been studied in the context of learning and memory [@zhu_prenatal_2004; @crino_corticosterone_2014-learn; @bebus_associative_2016; @abayarathna_effects_2020], neglecting how other important cognitive domains might be affected [but see @burger_antipredator_1998; @vila_pouca_quantity_2019]. In this vein, perception is an often overlooked, but fundamental cognitive process, involved in behaviours critical to survival and reproductive success such as locating food, avoiding predators, or interacting with conspecifics [@shettleworth; @hairston1982fish; @pyke1984optimal; @burger_antipredator_1998; @desfilis2003stimulus; @recio2023conspecific]. Understanding how prenatal conditions affect the ability to detect different stimuli can be fundamental to understanding the broader consequences of early-life environments on fitness.

Prenatal environments influence cognitive abilities because the brain is particularly sensitive during early stages of development [@zhu_prenatal_2004]. Early-life conditions can shape cognition by altering gene expression [@zhou2020effects], neurotransmitter production and neuroendocrine pathways [@amani2021perinatal], or brain structure and function [@amiel_effects_2017]. Additionally, developmental environments can have long-lasting effects on energy production and oxidative stress in the brain, which may be a key mechanism influencing cognitive function [@siegel1994basic; @du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018]. First, given the high energetic demands of cognitive processes [@mcnay_decreases_2000; @mann_coupling_2021; @alexandrov_neuronal_2022], cognitive performance is intrinsically linked to efficient mitochondrial respiration. Likewise, high cognitive abilities are typically associated with increased neuron density and functionality [@lefebvre_taxonomic_2011; @amiel_effects_2017], both of which can be impaired by excessive reactive oxygen species (ROS) production and oxidative stress [@finkel_oxidants_2000; @zhu_prenatal_2004; @du_dynamic_2009; @gong_chronic_2011; @hoffmann_mitochondrion_2018]. For instance, Hara et al. (2014) found that performance in visuospatial working-memory tasks is negatively correlated to the number of round-shaped mitochondria-associated with higher oxidative stress-per presynaptic terminal. Thus, understanding how early-life conditions affect mitochondrial physiology may provide insights into the mechanisms underlying cognitive performance.

Mitochondria are highly responsive to environmental conditions during early development, with potential short-term and long-lasting consequences [reviewed in @gyllenhammer2020developmental]. For example, environmental conditions or hormone exposure can cause immediate disruptions to metabolic function, triggering cascading effects on oxidative stress that may persist later in life [reviewed in @gyllenhammer2020developmental]. Among the range of environmental factors that can shape mitochondrial physiology, some exert particularly strong and consistent effects. Stress-related hormones and temperature are particularly relevant for mitochondrial function due to their strong effects on metabolic processes [@sapolsky_how_2000; @stier2022experimental; @crino2024eggs]. Vertebrates respond to stressful situations through the activation of physiological processes including the production of  glucocorticoids (GCs), which are also important metabolic regulators [@sapolsky_how_2000]. In ectothermic vertebrates, temperature plays a similar role in regulating metabolic activity [@stier2022experimental; @crino2024eggs]. In fact, heightened GC levels or changes in the thermal environment during early development have profound and sustained effects on mitochondrial respiration efficiency or oxidative stress [@zhu_prenatal_2004; @song2009prenatal; @costantini2011meta; @haussmann_embryonic_2012; @treidel2016temperature; @stier2022experimental; @crino2024eggs]. However, how the early environment influences mitochondrial physiology and its associated cognitive consequences remains largely unknown outside of mammals [see @chaudhari2022early for a review]. In addition, whether and how these interactions affect perceptual abilities, such as prey detection, is still largely unexplored.

@@

#### Animal husbandry  
*Breeding colony* -- The lizards tested came from a breeding colony established in the laboratory in 2019. The colony consisted of 270 adults of *L. delicata* housed in plastic containers (41.5 L x 30.5 W x 21 H cm) with six lizards (two males and four females) per enclosure. Enclosures were provided with non-stick matting, shelter, and several small water dishes. Water was given daily, and lizards were fed approx. 40 mid-size crickets (_Acheta domestica_) per enclosure three days a week. Crickets were dusted with calcium weekly and multivitamin and calcium biweekly. Room temperatures were set to 22-24 ºC, but enclosures were also provided with a heat chord and a heat lamp following a 12 h light:12 h dark cycle keeping warm side of enclosures is usually at 34 ºC.

*Egg collection and incubation* -- Between mid-October 2022 and the end of February 2023, we provide females with a place to lay the eggs by placing a small box (12.5 L x 8.3 W x 5 H cm) with moist vermiculite in one side of the communal enclosures. These boxes were checked three days a week for eggs. After collection, eggs were treated with either CORT or a vehicle control (see CORT and temperature manipulation below) and placed in individual cups (80 mL) with moist vermiculite (12 parts water to 4 parts vermiculite). Cups were covered with cling wrap to retain moisture and left at one of two different incubation temperatures (see CORT and temperature manipulation below) until hatching.

*Hatchlings* -- Incubators were checked three times a week for hatchlings. Hatched lizards were placed in individual enclosures in individual enclosures (18.7L x 13.2W x 6.3H cm) provided with nonstick matting and a small water dish until the beginning of the experiment. During this period, lizards were given water daily and received 3-6 small *A. domestica* crickets three times a week. All care otherwise follows similar protocols to adults (see above).

@@

To test the interactive effects of CORT and incubation temperature, we manipulated CORT concentrations in eggs and incubated them under one of two temperature regimes (see @fig-Methods A). We used a partial split clutch design where eggs from a given clutch were distributed equally across the four treatments when clutch sizes were larger than four and randomly across treatments when less than four. Eggs were topically supplied with either: a) 5 µL of crystalline corticosterone (Sigma, Cat. No. C2505) dissolved in 100% ethanol at a final 10 pg CORT/mL concentration (CORT treatment), or b) an equal volume of 100% Ethanol (Control treatment). We selected doses based on previous studies where CORT treatment increased mean yolk CORT levels by approximately 2 standard deviations above the mean natural concentration [@crino2024eggs]. Eggs were then incubated at either cold  (23 ± 3 ºC) or hot (28 ± 3 ºC) (see @fig-Methods). These temperatures are within the natural limits in _L. delicata_ [@cheetham2011embryonic].
